ForOpenCL: Transformations Exploiting Array Syntax in Fortran for Accelerator Programming

### 2011-12-12 papers/IJCSE/response_to_reviews.txt

Dear Reviewers-

This paper is an extended and updated version of the paper by the same title
that was submitted to the GPUScA 2011 workshop.  The first round of reviews
that were for the workshop resulted in us making a set of changes to add
detail in addressing the scope of work that the paper does and does not
cover.  Specifically, we state that the transformation technique descriped
in the paper is focused on the pure elementatal functions that are
turned into OpenCL kernels, and do not perform loop or other analyses that
are used by other methods to infer what loop bodies should be turned into
kernels.  We also provided updated information to reflect the most
recent Fortran 2008 standard treatment of pure and elemental functions.

The revised version of the GPUScA paper contains the following changes in
addition to those made based on the reviewer comments from the workshop:

- Added a subsection that details the ROSE-based transformation that does
  the actual Fortran to OpenCL conversion of the kernel.  This was not
  present in the workshop paper due to space constraints.  Based on
  comments during the presentation of the paper and discussions with
  readers, we believe that this makes the paper more complete.

- We added an additional figure to illustrate the halo concept to augment
  the discussion in the text.  This was again based on feedback that we
  received from readers in order to provide clarity with the additional
  space in the journal submission.

- Updated reference set to include recent activity in similar efforts,
  specifically the OpenACC work that was presented at Supercomputing 2011.

- Revision of the CFL condition presented in the shallow water model
  to address a missing term that, while unrelated to the transformation,
  is important for correctly describing the shallow water example.

- Numerous tweaks and minor text edits for clarification and presentation.

- Editing to formatting to match the LaTeX style for the journal.

Thank you for your time to review this updated version of the paper.

### 2011-12-12 papers/IJCSE/rose-transform.tex
final edits from craig

Programmers using ForOpenCL write kernel code in standard Fortran, and
a sequence of code transformations are applied to generate equivalent
code in C-based OpenCL.  These transformations are fairly
straightforward, and map a restricted subset of Fortran directly to
the C equivalent.  The transformation is based on a regular traversal
of the ROSE Sage AST that represents the Fortran code.  The visitor
used in the traversal handles four types of Sage AST Nodes: {\tt
  SgAllocateStatment}, {\tt SgFunctionDeclaration}, {\tt
  SgVariableDeclaration}, and {\tt SgExprStatement}.  These encompass
the primary activities that are commonly encountered in computational
kernels: allocation, variable and function declaration, and
expressions that correspond to arithmetic operations.  Allocate
statements are mapped to local tile variables in the kernel.  The
function declarations for the pure elemental Fortran functions are
mapped to OpenCL kernel function declarations.  Variable declarations and
expressions based on primitive (non-record) numerical types are mapped
to their direct equivalent on the C side.

@@

In order to perform the language conversion, two ROSE {\tt SgSourceFile}
objects are created -- one corresponding to the input Fortran code, and
one corresponding to the output OpenCL code.  The transformation performs
a traversal of the input Fortran code during which AST nodes are inserted
into the OpenCL code to represent the equivalent code in the output language.
The input Fortran code is the unmodified code that defines the pure
elemental functions and related code that represent the computations to be
performed on the accelerator.  The output is seeded with a template C
source file ({\tt cl\_template.c}) that comes pre-defined with a set of
helper functions that are used by the generated OpenCL code.  These
generic pre-defined functions are primarily related to memory indexing (such
as mapping 2D logical array indices to physical 1D addresses).  Our current
prototyle includes a limited set of helper functions sufficient to demonstrate
the 2D shallow water model code, but could conceivably be expanded to include
higher dimensional arrays with little effort.

>>>>

Programmers using ForOpenCL write kernel code in standard Fortran, and
a sequence of code transformations are applied to generate equivalent
code in C-based OpenCL.  These transformations are fairly
straightforward, and map a restricted subset of Fortran directly to
the C equivalent.  The transformation is based on a regular traversal
of the ROSE Sage AST that represents the Fortran code.  The visitor
used in the traversal handles four types of Sage AST Nodes: {\tt
  SgAllocateStatement}, {\tt SgFunctionDeclaration}, {\tt
  SgVariableDeclaration}, and {\tt SgExprStatement}.  These encompass
the primary activities that are commonly encountered in computational
kernels: allocation, variable and function declaration, and
expressions that correspond to arithmetic operations.  Allocate
statements are mapped to local tile variables in the kernel.  The
function declarations for the pure elemental Fortran functions are
mapped to OpenCL kernel function declarations.  Variable declarations and
expressions based on primitive (non-record) numerical types are mapped
to their direct equivalent on the C side.

@@

In order to perform the language conversion, two ROSE {\tt SgSourceFile}
objects are created -- one corresponding to the input Fortran code, and
one corresponding to the output OpenCL code.  The transformation performs
a traversal of the input Fortran code during which AST nodes are inserted
into the OpenCL code to represent the equivalent code in the output language.
The input Fortran code is the unmodified code that defines the pure
elemental functions and related code that represent the computations to be
performed on the accelerator.  The output is seeded with a template C
source file ({\tt cl\_template.c}) that comes pre-defined with a set of
helper functions that are used by the generated OpenCL code.  These
generic pre-defined functions are primarily related to memory indexing (such
as mapping 2D logical array indices to physical 1D addresses).  Our current
prototype includes a limited set of helper functions sufficient to demonstrate
the 2D shallow water model code, but could conceivably be expanded to include
higher dimensional arrays with little effort.
